Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model

### 2023-03-09 src/docs_bayescomp-poster/bayescomp-poster.Rmd
Almost finished with the poster

---
main_topsize: 0.08
main_bottomsize: 0.03
poster_height: "36in"
poster_width: "48in"
font_family: "Open Sans"
# Essentials
title: Integrated nested Laplace approximations for extended latent Gaussian models with application to the Naomi HIV model
author:
  - name: "Adam Howes"
    affil: 1, 2
    main: true
    twitter: adamhowes
    email: ath19@ic.ac.uk
  - name: "Alex Stringer"
    affil: 3
  - name: "Seth R. Flaxman"
    affil: 4
  - name: "Jeffrey W. Eaton"
    affil: 2
affiliation:
  - num: 1
    address: Department of Mathematics, Imperial College London
  - num: 2
    address: MRC Centre for Global Infectious Disease Analysis, School of Public Health, Imperial College London
  - num: 3
    address: Department of Statistics and Actuarial Science, University of Waterloo
  - num: 4
    address: Department of Computer Science, University of Oxford
logoleft_name: "IMP_ML_W_N.png"
logoright_name: "UW_ML_W_N.png"
# Style & formatting
column_numbers: 3
title_textsize: "65px"
author_textsize: "45px"
authorextra_textsize: "35px"
affiliation_textsize: "25px"
caption_fontsize: "20px"
# Title box
title_textcol: "#ffffff"
author_textcol: "#ffffff"
affiliation_textcol: "#ffffff"
# Main
main_fontfamily: "Open Sans"
# Poster body options
primary_colour: '#3b77D8'
secondary_colour: '#3b77D8'
accent_colour: "#5d19c4"
body_bgcol: "#ffffff"
body_textsize: "35px"
body_textcol: "#000000"
reference_textsize: "25px"
sectitle_textsize: "40px"
# Standard options
bibliography: citations.bib
output: 
  posterdown::posterdown_html:
    self_contained: false
    pandoc_args: --mathjax
    highlight: espresso
    number_sections: false
---

@@

* Approximate Bayesian inference method using adaptive Gauss-Hermite quadrature and Laplace approximation
* Implemented as a part of the `aghq` package, allowing easy use for any model with a `TMB` C++ user template
* Motivated by a challenging problem in evidence synthesis for small-area estimation of HIV

@@

* Spatio-temporal small-area estimation model of HIV indicators which combines data from household surveys, antenatal care clincs, and routine service provision of antiretroviral therapy [@eaton2021naomi]
* Yearly estimation process where the model is run interactively by country teams using the web-app `naomi.unaids.org`
* Inference currently conducted with empirical Bayes and a Gaussian approximation to the latent field using Template Model Builder `TMB` [@kristensen2016tmb]
* Model incompatible with the `R-INLA` implementation of integrated nested Laplace approximations [@rue2009approximate] and takes days to get accurate answers with Markov chain Monte Carlo via `tmbstan` [@monnahan2018no]
* Fits into the recently described class of extended latent Gaussian models [@stringer2022fast]

@@

### Laplace approximation

* Integrate out variables using a Gaussian approximation in the denominator
$$
p(\theta, y) \approx \tilde p_\text{LA}(\theta, y)
$$
where

### Adaptive Gauss-Hermite Quadrature

* Approximate integrals by $\int_\Theta p(\theta) \text{d} \theta \approx \lvert L \rvert \sum_{z \in \mathcal{Q}(m, k)} p(\hat \theta + Lz) \omega(z)$ with Gauss-Hermite quadrature rule $z \in \mathcal{Q}(m, k)$ adapted based upon the mode $\hat \theta = \text{argmax}_{\theta \in \Theta} \in p(\theta)$ and lower Cholesky $LL^\top = - \partial_\theta^2 \log p(\theta) \rvert_{\theta = \hat \theta}$ of the target

### Our algorithm

Given C++ user template for $- \log p(y, x, \theta)$:

```{r echo=FALSE, fig.align='center', out.width='100%'}
knitr::include_graphics("algorithm-flowchart.png")
```

## Inference comparison

* Compare posterior inferences from `TMB`, `aghq` and `tmbstan` using Kolmogorov-Smirnov tests on posterior marginals
* Find that...

<font size="6"> **Funding** AH was supported by the EPSRC and Bill & Melinda Gates Foundation. This research was supported by the MRC Centre for Global Infectious Disease Analysis. </font>

@@

<font size="6"> <div id="refs" class="references"></div> </font>

>>>>

---
poster_height: "46.8in"
poster_width: "33.1in"
font_family: "Open Sans"
# Essentials
title: Fast approximate Bayesian inference for small-area estimation of HIV indicators using the Naomi model
# titlebox_height: "10%"
author:
  - name: "Adam Howes"
    affil: 1, 2
    main: true
    email: ath19@ic.ac.uk
  - name: "Alex Stringer"
    affil: 3
  - name: "Seth R. Flaxman"
    affil: 4
  - name: "Jeffrey W. Eaton"
    affil: 2
affiliation:
  - num: 1
    address: Department of Mathematics, Imperial College London
  - num: 2
    address: MRC Centre for Global Infectious Disease Analysis, School of Public Health, Imperial College London
  - num: 3
    address: Department of Statistics and Actuarial Science, University of Waterloo
  - num: 4
    address: Department of Computer Science, University of Oxford
logoleft_name: "mrc-gida-icl-uw-uo.png"
# Style & formatting
column_numbers: 3
title_textsize: "75px"
author_textsize: "55px"
authorextra_textsize: "35px"
affiliation_textsize: "35px"
caption_fontsize: "20px"
# Title box
title_textcol: "#000000"
author_textcol: "#000000"
affiliation_textcol: "#000000"
# Main
main_fontfamily: "Open Sans"
# Poster body options
body_bgcol: "#ffffff"
body_textsize: "36px"
body_textcol: "#000000"
reference_textsize: "25px"
sectitle_textsize: "40px"
link_col: "#000000"
# Standard options
bibliography: citations.bib
output: 
  posterdown::posterdown_html:
    self_contained: false
    pandoc_args: --mathjax
    highlight: espresso
    number_sections: false
---

```{css, echo=FALSE}
div.logo_left{
  width: 35%;
  background-color: #E5E5E5;
}
div.poster_title{
  width: 60%;
}
div.title_container{
  background-color: #E5E5E5;
}
code{
  background-color: #ffffff;
  color: #000000;
}
.poster_body{
  column-rule-color: #ffffff;
}
```

@@

* Approximate Bayesian inference method using Laplace approximations and adaptive Gauss-Hermite quadrature 
* Motivated by an evidence synthesis model for small-area estimation of HIV indicators in sub-Saharan Africa
* Implemented as a part of the `aghq` package [@stringer2021implementing], allowing flexible use of the method for any model with a `TMB` C++ user template

@@

* District-level model of HIV indicators [@eaton2021naomi] which synthesises data from household surveys, antenatal care (ANC) clincs, and routine service provision of antiretroviral therapy (ART)
  * Combining evidence from multiple data sources helps overcome the limitations of any one
* Yearly estimation process: model run interactively by country teams using a web-app
  * Figure \@ref(fig:fig1) illustrates the seven stages of using the app
* Inference conducted in minutes using empirical Bayes (EB) and a Gaussian approximation via Template Model Builder `TMB` [@kristensen2016tmb]
* It would take days to get accurate answers with MCMC via `tmbstan` [@monnahan2018no], and this is not practical in this setting
* Motivates looking for a fast, approximate approach, that takes uncertainty in hyperparameters into account

```{r fig1, echo=FALSE, fig.align='center', out.width='100%', fig.cap='Model fitting occurs interactively in stages, taken from `naomi.unaids.org`'}
knitr::include_graphics("fig1.png")
```

## Extended latent Gaussian models

* Latent Gaussian model (LGM) [@rue2009approximate]: three stage hierarchical model with observations $y$, Gaussian latent field $x$ and hyperparameters $\theta$
* In an LGM the conditional mean depends on exactly one structured additive predictor $\mu_i = g(\eta_i)$ with $g: \mathbb{R} \to \mathbb{R}$
  * The `R-INLA` implementation of integrated nested Laplace approximations applies only to LGMs
* Extended latent Gaussian models (ELGM) remove this requirement such that $\mu_i = g(\eta_{\mathcal{J_i}})$ where $g_i: \mathbb{R}^{|\mathcal{J_i}|} \to \mathbb{R}$
* Naomi is an ELGM, not an LGM, because we model:
  * ANC indicators offset from household survey indicators
  * Incidence depending on prevalence and ART coverage
  * Observed data as aggregated finer processes
  * Individuals as possibly attending ART clinics outside their home district
* ELGM precision matrices less sparse than those of LGMs
* We extend work of @stringer2022fast in this setting to the challenging Naomi ELGM

@@

* Laplace approximation: Integrate out variables using a Gaussian approximation to the denominator
$$
p(\theta, y) \approx \tilde p_\text{LA}(\theta, y) = \frac{p(y, x, \theta)}{\tilde p_\text{G}(x \, | \, \theta, y)} \Big\rvert_{x = \hat x(\theta)}
$$
where $\tilde p_\text{G}(x \, | \, \theta, y) = \mathcal{N}(x \, | \, \hat x(\theta), \mathbf{H}(\theta)^{-1})$
  * Use automatic differentiation via `CppAD` in `TMB`
* Adaptive Gauss-Hermite Quadrature:
$$
\int_\Theta p(\theta) \text{d} \theta \approx \lvert L \rvert \sum_{z \in \mathcal{Q}(m, k)} p(\hat \theta + Lz) \omega(z)
$$
where the Gauss-Hermite quadrature rule $z \in \mathcal{Q}(\dim(\theta), k)$ is adapted based upon the mode $\hat \theta = \text{argmax}_{\theta \in \Theta} p(\theta)$ and lower Cholesky $LL^\top = - \partial_\theta^2 \log p(\theta) \rvert_{\theta = \hat \theta}$ of the target
  * $k$ is the number of quadrature points per dimension
* Algorithm (called `adam` for now) summarized by Figure \@ref(fig:fig2)

```{tikz, fig2, echo=FALSE, fig.align='center', fig.cap = "Flowchart describing the algorithm", fig.ext = 'png', cache=TRUE}
\usetikzlibrary{arrows}
\begin{tikzpicture}[node distance=2.5cm, auto,>=latex', thick, scale = 1]
\node (1) {$p(\theta, x, y)$};
\node (2) [below of=1] {$\tilde p_\text{LA}(x_i, \theta, y)$};
\node (3) [below of=2] {$\tilde p_\text{LA}(x_i, y)$};
\node (4) [below of=3] {$\tilde p_\text{AQ}(x_i \, | \, y)$};
\node (5) [below of=4] {Output};

\node (6) [right of=1, xshift=1cm] {$\tilde p_\text{LA}(\theta, y)$};
\node (7) [below of=6] {$\tilde p_\text{AQ}(y)$};
\node (8) [below of=7] {$\tilde p_\text{AQ}(\theta \, | \, y)$};
\node (9) [below of=8] {$\tilde p_\text{AQ}(\theta_j \, | \, y)$};
\node (10) [below of=9] {Output};

\draw[->] (1) to node {Laplace} (2);
\draw[->] (2) to node {AGHQ} (3);
\draw[->] (3) to node {Normalise} (4);
\draw[->] (4) to node {Spline} (5);
\draw[->] (1) to node {Laplace} (6);
\draw[->] (6) to node {AGHQ} (7);
\draw[->] (7) to node {Normalise} (8);
\draw[->] (8) to node {Marginalise} (9);
\draw[->] (9) to node {Spline} (10);
\end{tikzpicture}
```

## Application to Malawi data

```{r fig3, echo=FALSE, fig.align='center', out.width='100%', fig.cap='Model outputs for adults 15-49 in January 2016.'}
knitr::include_graphics("fig3.png")
```

* Relatively small country but still a large model
  * $\text{dim}(x)$ is 491 and $\text{dim}(\theta)$ is 24
* Fit inference methods `TMB`, `aghq`, `adam` and `tmbstan`
* Figure \@ref(fig:fig3) illustrates example model outputs: HIV prevalence, ART coverage, HIV incidence, and number of new infections, at the district level

```{r fig4, echo=FALSE, fig.align='center', out.width='100%', fig.cap='Inference reults and ECDF comparison for one $x_i$.'}
knitr::include_graphics("fig4.png")
```

* Compare inferences using Kolmogorov-Smirnov (KS) tests on hyperparameter, latent field, and output posterior marginals
* Figure \@ref(fig:fig4) illustrates the ECDF differences for one particular node in the model with $\text{KS}(\texttt{TMB}, \texttt{tmbstan}) = 0.09$ and $\text{KS}(\texttt{adam}, \texttt{tmbstan}) = 0.02$

## Future directions

* Scaling up the hyperparameter grid beyond EB, when any dense grid would be impractical ($k^{24}$ nodes)
  * Sparse grids
  * Dense grids on a subspace
* Add Laplace matrix algebra approximations [@wood2020simplified]
* More comprehensive inference comparisons
  * Maximum mean discrepancy
  * Pareto-smoothed importance sampling

**Interested?** Working notebooks and R code available from `github.com/athowes/elgm-inf`, or get in touch:

* `r fontawesome::fa("home", fill = "black")` `athowes.github.io`
* `r fontawesome::fa("envelope", fill = "black")` `ath19@ic.ac.uk`
* `r fontawesome::fa("twitter", fill = "black")` `adamhowes`

**Funding** AH was supported by the EPSRC and Bill & Melinda Gates Foundation. This research was supported by the MRC Centre for Global Infectious Disease Analysis.

@@

<font size="5"> <div id="refs" class="references"></div> </font>
